---
title: "Advent of Code 2019"
output: html_notebook
---

# Day 05

## load required libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gtools)
```

## load input data

```{r echo=TRUE, message=FALSE, warning=FALSE}
day05_input <- read_lines ("inputs/day05_01_input.txt") %>% strsplit(split = ',') %>% unlist() %>% as.integer()

sprintf('loaded instruction_stream: %s', paste(day05_input, collapse=','))
```

## part 01: write program function

> The TEST diagnostic program will run on your existing Intcode computer after a few modifications:

For reference, the following is exactly the "computer" written for the day 02 solution:

```{r echo=TRUE, message=FALSE, warning=FALSE}
old_intcode_computer <- function(program_stream, verbose = T) {
  i <- 1
  while((opcode <- program_stream[i]) != 99) {
    input1.adress <- program_stream[i+1]+1
    input2.adress <- program_stream[i+2]+1
    output.address <- program_stream[i+3]+1
    
    input1.value <- program_stream[input1.adress]
    input2.value <- program_stream[input2.adress]
      
    if(opcode == 1) {
      if(verbose) print(sprintf('instruction %s: writing %s + %s to position %s', (i+3)/4, input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value + input2.value
      
    } else if(opcode == 2) {
      if(verbose) print(sprintf('instruction %s: writing %s * %s to position %s', (i+3)/4, input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value * input2.value
    
    } else {
      stop(sprintf('unknown opcode %s', opcode))
    }
    
    i <- i+4
  }
  if(verbose) print(sprintf('exited gracefully, finding a opcode 99 at position %s', i))
  return(program_stream)
}
```

Now, this hs to be extended wth two new instructions:

- Opcode 3 takes a single integer as input and saves it to the position given by its only parameter.  
For example, the instruction 3,50 would take an input value and store it at address 50.  
- Opcode 4 outputs the value of its only parameter. For example, the instruction 4,50 would output the value at address 50.
- Second, you'll need to add support for parameter modes:
-- Parameter mode 0, position mode, which causes the parameter to be interpreted as a position  
-- Parameters in mode 1, immediate mode. In immediate mode, a parameter is interpreted as a value
-- Parameters that an instruction writes to will never be in immediate mode.

```{r}
get_opcode <- function(input) {
  input %>% as.character %>% strsplit(split='') %>% unlist %>% tail(2) %>% paste(collapse='') %>% as.integer
}
get_param_modes <- function(input) {
  input %>% as.character %>% strsplit(split='') %>% unlist %>% head(-2) %>% rev
}

intcode_computer.v1 <- function(program_stream, default.input = 1, verbose = T) {
  output <- c()
  
  i <- 1
  while((opcode <- get_opcode(program_stream[i])) != 99) {
    param_modes <- get_param_modes(program_stream[i])
      
    if(opcode == 1) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      output.address <- program_stream[i+3]+1
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      
      if(verbose) print(sprintf('writing %s + %s to position %s', input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value + input2.value
      i <- i+4
      
    } else if(opcode == 2) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      output.address <- program_stream[i+3]+1
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      
      if(verbose) print(sprintf('writing %s * %s to position %s', input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value * input2.value
      i <- i+4
      
    } else if(opcode == 3) {
      input <- default.input # for the ease of operation, commenting this out. gtools::ask(msg='Please provide input') %>% as.integer()
      
      write.address <- program_stream[i+1]+1
      
      if(verbose) print(sprintf('collected input %s, writing it to address %s', input, write.address))
      program_stream[write.address] <- input
      i <- i+2
    
    } else if(opcode == 4) {
      input.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input.value <- program_stream[input.adress]
      
      if(verbose) print(sprintf('instruction: writing %s to output', input.value))
      output <- c(output, input.value)
      
      if(input.value != 0 && program_stream[i+2] != 99) {
        stop(sprintf('non-0 output: %s. previously executed operation at address %s', input.value, i-1))
      }
      i <- i+2
      
    } else {
      stop(sprintf('unknown opcode %s. previously executed operation at address %s', opcode, i-1))
    }
    
  }
  
  if(verbose) print(sprintf('exited gracefully, finding a opcode 99 at position %s', i))
  return(output)
}
```

The validation actually comes as part of the execution, so we will run this diagnostic program.  
If all outputs are 0, except the last one before hitting the 99-code, then we are good.

```{r}
diagnostic_output <- intcode_computer.v1(day05_input, default.input = 1, verbose = T)
```

```{r}
sprintf('final diagnostic code: %s', diagnostic_output %>% tail(1))
```

## part 02: add some more instructions

Surprisingly, part 02 asks to add even more functionality to the computer...

- Opcode 5 is jump-if-true: if the first parameter is non-zero, it sets the instruction pointer to the value from the second parameter. Otherwise, it does nothing.  
- Opcode 6 is jump-if-false: if the first parameter is zero, it sets the instruction pointer to the value from the second parameter. Otherwise, it does nothing.  
- Opcode 7 is less than: if the first parameter is less than the second parameter, it stores 1 in the position given by the third parameter. Otherwise, it stores 0.  
- Opcode 8 is equals: if the first parameter is equal to the second parameter, it stores 1 in the position given by the third parameter. Otherwise, it stores 0.

```{r}
intcode_computer.v2 <- function(program_stream, default.input = 1, verbose = T, diagnostic_mode = F) {
  output <- c()
  
  i <- 1
  while((opcode <- get_opcode(program_stream[i])) != 99) {
    param_modes <- get_param_modes(program_stream[i])
      
    if(opcode == 1) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      output.address <- program_stream[i+3]+1
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      
      if(verbose) print(sprintf('writing %s + %s to position %s', input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value + input2.value
      i <- i+4
      
    } else if(opcode == 2) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      output.address <- program_stream[i+3]+1
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      
      if(verbose) print(sprintf('writing %s * %s to position %s', input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value * input2.value
      i <- i+4
      
    } else if(opcode == 3) {
      input <- default.input # for the ease of operation, commenting this out. gtools::ask(msg='Please provide input') %>% as.integer()
      
      write.address <- program_stream[i+1]+1
      
      if(verbose) print(sprintf('collected input %s, writing it to address %s', input, write.address))
      program_stream[write.address] <- input
      i <- i+2
    
    } else if(opcode == 4) {
      input.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input.value <- program_stream[input.adress]
      
      if(verbose) print(sprintf('instruction: writing %s to output', input.value))
      output <- c(output, input.value)
      
      # it seems that this part of the code is not required anymore - will make it dependent on diagnostic_mode parameter (default to F)
      if(diagnostic_mode & input.value != 0 && program_stream[i+2] != 99) {
        stop(sprintf('non-0 output: %s. previously executed operation at address %s', input.value, i-1))
      }
      i <- i+2
      
    } else if(opcode %in% c(5,6)) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      
      if(opcode == 5 & input1.value != 0) {
        i <- input2.value+1
      } else if(opcode == 6 & input1.value == 0) {
        i <- input2.value+1
      } else {
        i <- i+3
      }
      if(verbose) print(sprintf('jump: moving instruction pointer to address %s', i))
      
    } else if(opcode %in% c(7,8)) {
      input1.adress <- ifelse(is.na(param_modes[1]) | param_modes[1] == 0, program_stream[i+1]+1, i+1)
      input2.adress <- ifelse(is.na(param_modes[2]) | param_modes[2] == 0, program_stream[i+2]+1, i+2)
      output.address <- program_stream[i+3]+1
      
      input1.value <- program_stream[input1.adress]
      input2.value <- program_stream[input2.adress]
      output.value <- case_when(
        opcode == 7 ~ input1.value < input2.value,
        opcode == 8 ~ input1.value == input2.value,
        T ~ F
      ) %>% as.integer
      
      if(verbose && opcode == 7) print(sprintf('writing %s (%s < %s) to position %s', output.value, input1.value, input2.value, output.address))
      if(verbose && opcode == 8) print(sprintf('writing %s (%s == %s) to position %s', output.value, input1.value, input2.value, output.address))
      program_stream[output.address] <- output.value
      i <- i+4
      
    } else {
      stop(sprintf('unknown opcode %s. previously executed operation at address %s', opcode, i-1))
    }
    
  }
  
  if(verbose) print(sprintf('exited gracefully, finding a opcode 99 at position %s', i))
  return(output)
}
```


For validation, there are some examples given:

- `3,9,8,9,10,9,4,9,99,-1,8` - Using position mode, consider whether the input is equal to 8; output 1 (if it is) or 0 (if it is not).  

```{r}
intcode_computer.v2(c(3,9,8,9,10,9,4,9,99,-1,8), default.input = 9, verbose = F) == 0
```
```{r}
intcode_computer.v2(c(3,9,8,9,10,9,4,9,99,-1,8), default.input = 8, verbose = F) == 1
```

- `3,9,7,9,10,9,4,9,99,-1,8` - Using position mode, consider whether the input is less than 8; output 1 (if it is) or 0 (if it is not).

```{r}
intcode_computer.v2(c(3,9,7,9,10,9,4,9,99,-1,8), default.input = 9, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,9,7,9,10,9,4,9,99,-1,8), default.input = 8, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,9,7,9,10,9,4,9,99,-1,8), default.input = 7, verbose = F) == 1
```
 
- `3,3,1108,-1,8,3,4,3,99` - Using immediate mode, consider whether the input is equal to 8; output 1 (if it is) or 0 (if it is not). 

```{r}
intcode_computer.v2(c(3,3,1108,-1,8,3,4,3,99), default.input = 9, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,3,1108,-1,8,3,4,3,99), default.input = 8, verbose = F) == 1
```


- `3,3,1107,-1,8,3,4,3,99` - Using immediate mode, consider whether the input is less than 8; output 1 (if it is) or 0 (if it is not).

```{r}
intcode_computer.v2(c(3,3,1107,-1,8,3,4,3,99), default.input = 9, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,3,1107,-1,8,3,4,3,99), default.input = 8, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,3,1107,-1,8,3,4,3,99), default.input = 7, verbose = F) == 1
```

The part 02 task also provides some jump tests that take an input, then output 0 if the input was zero or 1 if the input was non-zero:

- `3,12,6,12,15,1,13,14,13,4,13,99,-1,0,1,9` (using position mode)

```{r}
intcode_computer.v2(c(3,12,6,12,15,1,13,14,13,4,13,99,-1,0,1,9), default.input = 0, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,12,6,12,15,1,13,14,13,4,13,99,-1,0,1,9), default.input = 1, verbose = F) == 1
```

- `3,3,1105,-1,9,1101,0,0,12,4,12,99,1` (using immediate mode)

```{r}
intcode_computer.v2(c(3,3,1105,-1,9,1101,0,0,12,4,12,99,1), default.input = 0, verbose = F) == 0
```

```{r}
intcode_computer.v2(c(3,3,1105,-1,9,1101,0,0,12,4,12,99,1), default.input = 1, verbose = F) == 1
```

And a final larger example:

- `3,21,1008,21,8,20,1005,20,22,107,8,21,20,1006,20,31,1106,0,36,98,0,0,1002,21,125,20,4,20,1105,1,46,104,999,1105,1,46,1101,1000,1,20,4,20,1105,1,46,98,99` - this example program uses an input instruction to ask for a single number.  
The program will then output 999 if the input value is below 8, output 1000 if the input value is equal to 8, or output 1001 if the input value is greater than 8.

```{r}

intcode_computer.v2(c(3,21,1008,21,8,20,1005,20,22,107,8,21,20,1006,20,31,1106,0,36,98,0,0,1002,21,125,20,4,20,1105,1,46,104,999,1105,1,46,1101,1000,1,20,4,20,1105,1,46,98,99), default.input = 7, verbose = F, diagnostic_mode = F) == 999
```

```{r}
intcode_computer.v2(c(3,21,1008,21,8,20,1005,20,22,107,8,21,20,1006,20,31,1106,0,36,98,0,0,1002,21,125,20,4,20,1105,1,46,104,999,1105,1,46,1101,1000,1,20,4,20,1105,1,46,98,99), default.input = 8, verbose = F, diagnostic_mode = F) == 1000
```

```{r}
intcode_computer.v2(c(3,21,1008,21,8,20,1005,20,22,107,8,21,20,1006,20,31,1106,0,36,98,0,0,1002,21,125,20,4,20,1105,1,46,104,999,1105,1,46,1101,1000,1,20,4,20,1105,1,46,98,99), default.input = 9, verbose = F, diagnostic_mode = F) == 1001
```

(all works well, but we had to remove the condition to test that non-zero outputs had to be followed by a 99-opcode.)  
  
So, finally, let's run this thing for the quiz input.  
This diagnostic test suite only outputs one number, the diagnostic code.

> What is the diagnostic code for system ID 5?

```{r}
diagnostic_code.v2 <- intcode_computer.v2(day05_input, default.input = 5, verbose = T, diagnostic_mode = F)
```

```{r}
sprintf('diagnosis program returned %s', diagnostic_code.v2)
```