---
title: "Advent of Code 2019"
output: html_notebook
---

# Day 02

## load required libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## load input data

```{r echo=TRUE, message=FALSE, warning=FALSE}
day02_input <- read_lines ("inputs/day02_01_input.txt") %>% strsplit(split = ',') %>% unlist() %>% as.integer()

sprintf('loaded instruction_stream: %s', paste(day02_input, collapse=','))
```

## part 01: write program function

> Opcode 1 adds together numbers read from two positions and stores the result in a third position.

> Opcode 2 works exactly like opcode 1, except it multiplies the two inputs instead of adding them.

> Once you're done processing an opcode, move to the next one by stepping forward 4 positions.

> 99 means that the program is finished and should immediately halt

```{r echo=TRUE, message=FALSE, warning=FALSE}
run_program <- function(program_stream, verbose = T) {
  i <- 1
  while((opcode <- program_stream[i]) != 99) {
    input1.adress <- program_stream[i+1]+1
    input2.adress <- program_stream[i+2]+1
    output.address <- program_stream[i+3]+1
    
    input1.value <- program_stream[input1.adress]
    input2.value <- program_stream[input2.adress]
      
    if(opcode == 1) {
      if(verbose) print(sprintf('instruction %s: writing %s + %s to position %s', (i+3)/4, input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value + input2.value
      
    } else if(opcode == 2) {
      if(verbose) print(sprintf('instruction %s: writing %s * %s to position %s', (i+3)/4, input1.value, input2.value, output.address))
      program_stream[output.address] <- input1.value * input2.value
    
    } else {
      stop(sprintf('unknown opcode %s', opcode))
    }
    
    i <- i+4
  }
  if(verbose) print(sprintf('exited gracefully, finding a opcode 99 at position %s', i))
  return(program_stream)
}
```

Then, validate the function on the examples given:

```{r}
validation_stream.input <- c(1,9,10,3,2,3,11,0,99,30,40,50)
validation_stream.output <- run_program(validation_stream.input, verbose = F)
identical(validation_stream.output, c(3500,9,10,70,2,3,11,0,99,30,40,50))
```

Looking good. Now for the mean trick:

> before running the program, replace position 1 with the value 12 and replace position 2 with the value 2.
> What value is left at position 0 after the program halts?

```{r}
day02_input.1202state <- day02_input
day02_input.1202state[2] <- 12
day02_input.1202state[3] <- 2

day02_final_state <- run_program(day02_input.1202state)

sprintf('final value at position 0: %s', day02_final_state[1])
```
("final value at position 0: 3562624")

## part 02: find the input to output 19690720

> In this program, the value placed in address 1 is called the noun, and the value placed in address 2 is called the verb.

> Each of the two input values will be between 0 and 99, inclusive.

> Each time you try a pair of inputs, make sure you first reset the computer's memory to the values in the program (your puzzle input) - in other words, don't reuse memory from a previous attempt.

```{r}
bruteforce_outputs <- function(program_stream, nouns = 0:99, verbs = 0:99, verbose = F) {
  runs <- tibble()
  for(noun in nouns) {
    for(verb in verbs) {
      run_state <- day02_input
      run_state[2] <- noun
      run_state[3] <- verb
      
      run <- tibble(
        noun = noun,
        verb = verb,
        initial_state = list(run_state),
        final_state = list(run_program(run_state, verbose = F)),
        output = final_state[[1]][1]
      )
      
      # don't use this.. too much output >_>
      if(verbose) print(sprintf('produced %s with noun: %s & verb: %s', run$output, noun, verb))
      
      runs <- runs %>%
        bind_rows(run)
    }
  }
  
  return(runs)
}
```

Again, validate the function by checking whether we can find the inputs to the first part of the puzzle, only based on its output.  
This should deliver values for a bunch of input combinations, where we should find the output 3562624 being produced by noun: 12 & verb: 2

```{r}
validation_io <- bruteforce_outputs(day02_input, nouns = 0:20, verbs = 0:20, verbose = F)
validation_io %>%
  filter(output == 3562624) %>%
  {.$noun == 12 & .$verb == 2} %>%
  all
```
This works, even if it may be a very dumb approach...  
Nobody said I can't be dumb.

```{r}
validation_io <- bruteforce_outputs(day02_input, nouns = 0:99, verbs = 0:99, verbose = F)
day02_result.input <- validation_io %>%
  filter(output == 19690720) %>%
  mutate(
    input_param = 100 * noun + verb
  )

sprintf('found exactly %s inputs that lead to the output 19690720. this would be the program(s) %s', nrow(day02_result.input), day02_result.input$input_param)
```
("found exactly 1 inputs that lead to the output 19690720. this would be the program(s) 8298")
