---
title: "Advent of Code 2019"
output: html_notebook
---

# Day 04

## load required libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## load input data

```{r echo=TRUE, message=FALSE, warning=FALSE}
day04_input <- tibble(
  pw = 145852:616942
)
```

## part 01: 

### function to extract valid passwords from input vector

This will return all passwords valid by the part01 definition:

> It is a six-digit number.
> The value is within the range given in your puzzle input.
> Two adjacent digits are the same (like 22 in 122345).
> Going from left to right, the digits never decrease; they only ever increase or stay the same (like 111123 or 135679).

```{r}
evaluate_pws.v1 <- function(pw_df, keep.details = F) {
  regex.double_digit <- sprintf('%s{2}',0:9) %>% paste(collapse='|')
  regex.decreasing_nums <- sprintf('%s[0-%s]', 1:9, 0:8) %>% paste(collapse='|')
  
  output <- pw_df %>%
    mutate(
      valid.length = nchar(pw) == 6,
      valid.double_digit = grepl(regex.double_digit, pw),
      valid.no_decreasing_nums = !grepl(regex.decreasing_nums, pw),
      valid = valid.length & valid.double_digit & valid.no_decreasing_nums
    )
  
  if(!keep.details) output <- output %>% select(-starts_with('valid.'))
  return(output)
}
```

Validate that with the given three pws provided:  
- 111111 meets these criteria (double 11, never decreases).  
- 223450 does not meet these criteria (decreasing pair of digits 50).  
- 123789 does not meet these criteria (no double).

```{r}
tibble(pw = c(111111, 223450, 123789)) %>%
  evaluate_pws.v1(keep.details = F)
```
Seems in line with the expectations.  
Now, running this for the input range.

```{r}
input_validation.v1 <- day04_input %>%
  evaluate_pws.v1(keep.details = T)

input_validation.v1 %>%
  filter(valid)
```
```{r}
input_validation.v1 %>%
  filter(valid) %>%
  nrow %>%
  sprintf('found %s possible pws for part 01 of the quiz', .)
```

## stricter pw validation function

This function will extend the v1 check by adding the extra criterion:  
> the two adjacent matching digits are not part of a larger group of matching digits.

```{r}
evaluate_pws.v2 <- function(pw_df, keep.details = F) {
  # regex.double_digit <- sprintf('%s{2}',0:9) %>% paste(collapse='|')
  regex.double_digit.strict <- sprintf('(?<!%s)%s{2}(?!%s)', 0:9, 0:9, 0:9) %>% paste(collapse='|')
  regex.decreasing_nums <- sprintf('%s[0-%s]', 1:9, 0:8) %>% paste(collapse='|')
  
  output <- pw_df %>%
    mutate(
      valid.length = nchar(pw) == 6,
      # valid.double_digit = grepl(regex.double_digit, pw),
      valid.double_digit.strict = grepl(regex.double_digit.strict, pw, perl = T),
      valid.no_decreasing_nums = !grepl(regex.decreasing_nums, pw),
      valid = valid.length & valid.double_digit.strict & valid.no_decreasing_nums
    )
  
  if(!keep.details) output <- output %>% select(-starts_with('valid.'))
  return(output)
}
```

Again, we validate this on the provided examples:  
- 112233 meets these criteria because the digits never decrease and all repeated digits are exactly two digits long.  
- 123444 no longer meets the criteria (the repeated 44 is part of a larger group of 444). 
- 111122 meets the criteria (even though 1 is repeated more than twice, it still contains a double 22).

```{r}
tibble(pw = c(112233, 123444, 111122)) %>%
  evaluate_pws.v2(keep.details = F)
```
This looks good.  
Now, run this new function on the input range.

```{r}
input_validation.v2 <- day04_input %>%
  evaluate_pws.v2(keep.details = T)

input_validation.v2 %>%
  filter(valid)
```

```{r}
input_validation.v2 %>%
  filter(valid) %>%
  nrow %>%
  sprintf('found %s possible pws for part 01 of the quiz', .)
```

