---
title: "Advent of Code 2019"
output: html_notebook
---

# Day 03

## load required libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## load input data

```{r echo=TRUE, message=FALSE, warning=FALSE}
day03_input <- read_lines("inputs/day03_01_input.txt") %>%
  strsplit(split=',')
```

## part 01: find closest wire-intersection

### function to lay out wires into dataframe

This function takes every wire-segment and lays it out into a table of `[x,y]`-coordinates.  
We can then later `summarise` this table easily to identify coordinates where more than one wire has been observed.

```{r}
lay_out_wires <- function(wire_list) {
  wires <- tibble()

  for(wire_i in 1:length(wire_list)) {
    wire <- tibble()
    
    segment <- 1
    x <- 0
    y <- 0
    len <- 0
    
    for(instruction in wire_list[[wire_i]]) {
      direction_i <- str_extract(instruction, '^[UDLR]')
      len_i <- str_extract(instruction, '(?<=^[UDLR])(\\d+$)') %>% as.integer()
      
      x_vector <- case_when(
        direction_i == 'R' ~ (x+1):(x+len_i),
        direction_i == 'L' ~ (x-1):(x-len_i),
        T ~ x:x
      )
      y_vector <- case_when(
        direction_i == 'U' ~ (y+1):(y+len_i),
        direction_i == 'D' ~ (y-1):(y-len_i),
        T ~ y:y
      )
      len_new <- len + len_i
      
      
      wire <- wire %>%
        bind_rows(
          tibble(
            wire = wire_i,
            x = x_vector,
            y = y_vector,
            len = (len+1):len_new,
            segment = segment,
            direction = direction_i
          )
        )
      
      segment <- segment + 1
      x <- x_vector %>% last
      y <- y_vector %>% last
      len <- len_new
    }
    
    wires <- wires %>%
      bind_rows(wire)
  }
  
  return(wires)
}

```


## function to find intersections within laid-out dataframe format

This function will summarise the produced full wire-layout by counting the distinct wire-id's observed at each `[x,y]` value-pair.  
When we filter for only those coordinates with `n > 1` distinct cable observations, this should give us the intersection points.

```{r}
find_intersections <- function(wire_df) {
  wire_df %>%
    group_by(x,y) %>%
    summarise(
      n_wires = n_distinct(wire)
    ) %>%
    filter(
      n_wires > 1
    ) %>%
    mutate(
      manhattan_dist_to_cport = x+y
    ) %>%
    arrange(manhattan_dist_to_cport)
}

```


## validate

```{r}
validation.wire_layout01 <- strsplit(c('R8,U5,L5,D3','U7,R6,D4,L4'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout01) %>%
  head(1) %>%
  pull(manhattan_dist_to_cport) %>%
  {. == 6}
```

```{r}
validation.wire_layout02 <- strsplit(c('R75,D30,R83,U83,L12,D49,R71,U7,L72','U62,R66,U55,R34,D71,R55,D58,R83'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout02) %>%
  head(1) %>%
  pull(manhattan_dist_to_cport) %>%
  {. == 159}

validation.wire_layout02 %>%
  group_by(x,y) %>%
  mutate(
    n_wires = n_distinct(wire)
  ) %>%
  filter(
    n_wires > 1
  ) %>%
  mutate(
    manhattan_dist_to_cport = x+y
  ) %>%
  arrange(manhattan_dist_to_cport)
  
```

```{r}
validation.wire_layout03 <- strsplit(c('R98,U47,R26,D63,R33,U87,L62,D20,R33,U53,R51','U98,R91,D20,R16,D67,R40,U7,R15,U6,R7'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout03) %>%
  head(1) %>%
  pull(manhattan_dist_to_cport) %>%
  {. == 135}
```


## find intersection of input

Now, run this on the provided input file.

```{r}
input_wires <- lay_out_wires(day03_input)

sprintf('laid out wire with length %s', input_wires %>% group_by(wire) %>% summarise(len = max(len)) %>% pull(len))
```
```{r}
input_wire_intersections <- find_intersections(input_wires)

input_wire_intersections
```

