---
title: "Advent of Code 2019"
output: html_notebook
---

# Day 03

## load required libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## load input data

```{r echo=TRUE, message=FALSE, warning=FALSE}
day03_input <- read_lines("inputs/day03_01_input.txt") %>%
  strsplit(split=',')
```

## part 01: find closest wire-intersection

### function to lay out wires into dataframe

This function takes every wire-segment and lays it out into a table of `[x,y]`-coordinates.  
We can then later `summarise` this table easily to identify coordinates where more than one wire has been observed.

```{r}
lay_out_wires <- function(wire_list) {
  wires <- tibble()

  for(wire_i in 1:length(wire_list)) {
    wire <- tibble()
    
    segment <- 1
    x <- 0
    y <- 0
    length <- 0
    
    for(instruction in wire_list[[wire_i]]) {
      direction_i <- str_extract(instruction, '^[UDLR]')
      length_i <- str_extract(instruction, '(?<=^[UDLR])(\\d+$)') %>% as.integer()
      
      x_vector <- case_when(
        direction_i == 'R' ~ (x+1):(x+length_i),
        direction_i == 'L' ~ (x-1):(x-length_i),
        T ~ x:x
      )
      y_vector <- case_when(
        direction_i == 'U' ~ (y+1):(y+length_i),
        direction_i == 'D' ~ (y-1):(y-length_i),
        T ~ y:y
      )
      length_new <- length + length_i
      
      
      wire <- wire %>%
        bind_rows(
          tibble(
            wire = wire_i,
            x = x_vector,
            y = y_vector,
            length = (length+1):length_new,
            segment = segment,
            direction = direction_i
          )
        )
      
      segment <- segment + 1
      x <- x_vector %>% last
      y <- y_vector %>% last
      length <- length_new
    }
    
    wires <- wires %>%
      bind_rows(wire)
  }
  
  return(wires)
}

```


## function to find intersections within laid-out dataframe format

This function will summarise the produced full wire-layout by counting the distinct wire-id's observed at each `[x,y]` value-pair.  
When we filter for only those coordinates with `n > 1` distinct cable observations, this should give us the intersection points.

```{r}
find_intersections <- function(wire_df) {
  wire_df %>%
    group_by(x,y) %>%
    mutate(
      n_wires = n_distinct(wire)
    ) %>%
    filter(
      n_wires > 1
    ) %>%
    group_by(x,y,wire) %>%
    mutate(
      wire.shortest_path_length = min(length)
    ) %>%
    group_by(x,y) %>%
    summarise(
      shortcircuit.manhattan_dist_to_central_port = first(abs(x)+abs(y)),
      shortcircuit.min_wire_length = min(length),
      shortcircuit.min_full_circuit_length = sum(wire.shortest_path_length)
    ) %>%
    arrange(shortcircuit.manhattan_dist_to_central_port, x, y)
}

```


Now, we validate this on the provided examples.

```{r}
validation.wire_layout01 <- strsplit(c('R8,U5,L5,D3','U7,R6,D4,L4'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout01) %>%
  head(1) %>%
  pull(shortcircuit.manhattan_dist_to_central_port) %>%
  {. == 6}
```

```{r}
validation.wire_layout02 <- strsplit(c('R75,D30,R83,U83,L12,D49,R71,U7,L72','U62,R66,U55,R34,D71,R55,D58,R83'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout02) %>%
  head(1) %>%
  pull(shortcircuit.manhattan_dist_to_central_port) %>%
  {. == 159}
```

```{r}
validation.wire_layout03 <- strsplit(c('R98,U47,R26,D63,R33,U87,L62,D20,R33,U53,R51','U98,R91,D20,R16,D67,R40,U7,R15,U6,R7'), split=',') %>% lay_out_wires
find_intersections(validation.wire_layout03) %>%
  head(1) %>%
  pull(shortcircuit.manhattan_dist_to_central_port) %>%
  {. == 135}
```
This looks good, so we can run this on the provided input file to get the result for the first part of the quiz.  
First, let's "lay out" the wires that we loaded from the input file.

```{r}
input_wires <- lay_out_wires(day03_input)

sprintf('laid out wire with length %s', input_wires %>% group_by(wire) %>% summarise(length = max(length)) %>% pull(length))
```
This being done, we can find all intersections of the two wires and take a look at the coordinates they meet at.

```{r}
input_wire_intersections <- find_intersections(input_wires)

input_wire_intersections
```

## part 02: find shortest circuit

Finding the shortest circuit-length has been covered by above summary function, but we should validate this as well.  
This would probably give weird results if we use more than 2 wires (because the `sum` would include three wire lengths), but that's not really asked for here...  

So, running this again for the three previously loaded examples:

```{r}
find_intersections(validation.wire_layout01) %>%
  arrange(shortcircuit.min_full_circuit_length) %>%
  head(1) %>%
  {.$shortcircuit.min_full_circuit_length == 30}
```

```{r}
find_intersections(validation.wire_layout02) %>%
  arrange(shortcircuit.min_full_circuit_length) %>%
  head(1) %>%
  {.$shortcircuit.min_full_circuit_length == 610}
```

```{r}
find_intersections(validation.wire_layout03) %>%
  arrange(shortcircuit.min_full_circuit_length) %>%
  head(1) %>%
  {.$shortcircuit.min_full_circuit_length == 410}
```

All worked out perfectly, so now to run this on the input and go to bed :)

```{r}
input_wire_intersections %>%
  arrange(shortcircuit.min_full_circuit_length) %>%
  head(1) %>%
  pull(shortcircuit.min_full_circuit_length) %>%
  sprintf('The shortest circuit closed by the wires has a total length of %s units.', .)
```

